---
title: "Crime Clustering in Chicago"
author: "Alan Zablocki"
date: "August 3, 2016"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Setup

We load the required libraries.
```{r}
library(raster)
library(sp)
library(maptools)
gpclibPermit()
library(maptools)
library(mapdata)
library(ggmap)
library(ggplot2)
library(geosphere)
library(maps)
library(animation)
```
Load the theft data for Chicago
```{r}
mydata  = read.csv("Chicago_crimes_2008_somethefts_only_HP.csv")
head(mydata)
```
Select only relevant columns
```{r}
mydata <- mydata[,c(2,3,5,8,9,10)]
colnames(mydata) <- c("iucr","offense","locat","year","lati","long")
```
Subset data, include street theft only
```{r}
mydata_STREET_THEFT <- mydata[mydata$locat == "STREET", ]
```
Remove any NAs
```{r}
mydata_cleaned <- na.omit(mydata_STREET_THEFT)
```
Select 2014 as an example
```{r}
theft14 <- mydata_cleaned[mydata_cleaned$year == "2014", ]
head(theft14)
dim(theft14)
```
Find the range for long/lat in this dataset
```{r}
min(abs(theft14[,6]))
min(abs(theft14[,5]))
max(abs(theft14[,6]))
max(abs(theft14[,5]))
```
```{r}
# Get a map of Chicago with appropriate zoom level
require(ggmap)
map_of_Chicago <- get_map(location = c(long = -87.60, lati = 41.80), zoom = 13, maptype = "roadmap", scale = 2)
```
For now we will leave out the highly zoomed in area around the University of Chicago
```r
small_plot <- ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_density2d(aes(x = long, y = lati, fill = ..level.. , alpha = ..level..),bins = 25, colour = "black") + coord_map(projection="mercator", xlim=c(-87.607,-87.585), ylim=c(41.785,41.803)) + guides(fill = FALSE,alpha=FALSE,size=FALSE) + geom_point(aes(x = long, y = lati), size = 1, fill = "red", shape = 21)
small_plot
```
```{r largerHP, echo = FALSE}
# larger area around HP
largerHP <- ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_density2d(aes(x = long, y = lati, fill = ..level.. , alpha = ..level..),bins = 25, colour = "black") + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82)) + guides(fill = FALSE,alpha=FALSE,size=FALSE) + geom_point(aes(x = long, y = lati), size = 1, fill = "red", shape = 21)
plot(largerHP)
```
```{r}
# Perform k-means clustering on the whole data set.
# Match x and y axis to long / lati
kmeans_data <- theft14[,6:5]
head(kmeans_data)
```
```{r}
# Find an appropriate number of clusters based on 'elbow' plot
crime0 <- na.omit(kmeans_data)
crime <- data.matrix(crime0)
str(crime)
cl <- kmeans(crime,5)
class(cl)
str(cl)

# helper functions to get the 'elbow' plot
kmeans.wss.k <- function(crime, k){
  km = kmeans(crime, k)
  return (km$tot.withinss)
}
kmeans.dis <- function(crime, maxk){
  dis=(nrow(crime)-1)*sum(apply(crime,2,var))
  dis[2:maxk]=sapply (2:maxk, kmeans.wss.k, crime=crime)
  return(dis)
}
```
```{r DIST1, echo=FALSE}
# Start with maximum of 20 clusters
maxk = 20
dis = kmeans.dis(crime, maxk);
DIST1 <- plot(1:maxk, dis, type='b', xlab="Number of Clusters",
     ylab="Distortion",
     col="blue")
DIST1
```
```{r}
# Seems like 10 clusters should do it. Let's animate this.
# animation is nice, but takes up a lot of the write up
#cl<- kmeans.ani(kmeans_data, 10)
```
```{r}
#Examine the cluster plot
#fit1 <- kmeans(crime,10)
#mydata1 <- data.frame(mydata,fit1$cluster)
#aggregate(crime,by=list(fit1$cluster),FUN=mean)
#clusplot(mydata1,fit1$cluster, color = TRUE, shade = TRUE, labels = 2, lines=0)
```
```{r}
# set random seed for reproducing results
# plotting clustered data coloured by cluster solution using k-means
set.seed(1234)
dfX <- data.frame(long=theft14$long,lati=theft14$lati)
km <- kmeans(dfX,centers=10)
kmc <- data.frame(long=km$centers[,1],lati=km$centers[,2])
dfX$clust <- km$cluster
ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_density2d(aes(x = long, y = lati, fill = ..level.. , alpha = ..level..),bins = 25, colour = "black") + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82)) + guides(fill = FALSE,alpha=FALSE,size=FALSE) + geom_point(aes(x = long, y = lati), size = 2, fill = "red", shape = 21) + geom_point(data=dfX, aes(x=long, y=lati, color=factor(clust)), size=2) + geom_point(data=kmc, aes(x=long, y=lati), size=2)

# What if we perform the clustering on distances instead of latitude/longitude pairs?

# helper function

geo.dist = function(df) {
  require(geosphere)
  d <- function(i,z){         # z[1:2] contain long, lat
    dist <- rep(0,nrow(z))
    dist[i:nrow(z)] <- distHaversine(z[i:nrow(z),1:2],z[i,1:2])
    return(dist)
  }
  dm <- do.call(cbind,lapply(1:nrow(df),d,df))
  return(as.dist(dm))
}

df_lat <- data.frame(long=theft14$long,lati=theft14$lati)
df_dist <- geo.dist(df_lat)
set.seed(1234)
km_dist <- kmeans(df_dist,centers = 10)
set.seed(1234)
km_lat <- kmeans(df_lat,centers = 10)
df_with_dist <- data.frame(long=theft14$long,lati=theft14$lati)
df_lat$clust <- km_lat$cluster
df_with_dist$clust <- km_dist$cluster
kmc_lat <- data.frame(long=km_lat$centers[,1],lati=km_lat$centers[,2])
kmc_dist <- data.frame(long=km_dist$centers[,1],lati=km_dist$centers[,2])

ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_point(data=df_with_dist, aes(x=long, y=lati, color=factor(clust)), size=3) + geom_point(aes(x = long, y = lati), size = 1, fill = "red", shape = 21) + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82))

ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_point(data=df_lat, aes(x=long, y=lati, color=factor(clust)), size=3) + geom_point(aes(x = long, y = lati), size = 1, fill = "red", shape = 21) + geom_point(data=kmc_lat, aes(x=long, y=lati), size=2, fill="black") + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82))

```

Performing k-means on crime data with longitude and latitude gives similar results to when we run k-means on distances, but some of the data is assigned to different clusters. 

Although, 10 clusters seems like enough, in reality we do care about a maximum distance that a cluster will cover. In the map, we see that each cluster covers a number of blocks, which is not ideal. 

The next steps are:

1 - Increase the number of clusters in k-means to cover smaller areas
2 - Perform hierarchical clustering and invetigate a useful upper bound for a cluster
3 - Look at different clustering methods like dbscan, pam, clara and their possible application to this problem.
4 - Analyze clusters for smaller area, immediately surrounding the HP UofC campus

```{r}
set.seed(1234)
dfX2 <- data.frame(long=theft14$long,lati=theft14$lati)
d <- geo.dist(dfX2)
hc <- hclust(d)
plot(hc) # plot dendrogram
dfX2$clust <- cutree(hc,k=10)
ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_point(data=dfX2, aes(x=long, y=lati, color=factor(clust)), size=3) + geom_point(aes(x = long, y = lati), size = 1, fill = "red", shape = 21) + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82))
```

```{r}
set.seed(1234)
dfX2 <- data.frame(long=theft14$long,lati=theft14$lati)
d <- geo.dist(dfX2)
hc <- hclust(d)
plot(hc) # plot dendrogram
dfX2$clust <- cutree(hc,k=40)
ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + geom_point(data=dfX2, aes(x=long, y=lati, color=factor(clust)), size=3) + geom_point(aes(x = long, y = lati), size = 1, fill = "red", shape = 21) + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82))
```
How does k-means perform for 40 clusters?
```{r}
set.seed(1234)
dfX <- data.frame(long=theft14$long,lati=theft14$lati)
km <- kmeans(dfX,centers=40)
kmc <- data.frame(long=km$centers[,1],lati=km$centers[,2])
dfX$clust <- km$cluster
ggmap(map_of_Chicago, extent = "normal", maprange=FALSE) %+% theft14 + aes(x = long, y = lati) + coord_map(projection="mercator", xlim=c(-87.63,-87.57), ylim=c(41.77,41.82)) + guides(fill = FALSE,alpha=FALSE,size=FALSE) + geom_point(data=dfX, aes(x=long, y=lati, color=factor(clust)), size=2) + geom_point(data=kmc, aes(x=long, y=lati), size=2)
```
